Powering Hidden Markov Model by Generative Models

### 2019-07-12 manuscript/main.tex
In E-step, take the summation out
correct typos
formulate solution to update to pi

Since model $\bm{H}$ contains hidden sequential variable $\ubm{s}$, we can not directly solve the maximum likelihood problem in \autoref{eq:ml-of-hmm}. We use expectation maximization (EM) to address the hidden variable problem by
\begin{itemize}
\item E-step: % the posterior probability of $\ubm{s}$:
  % \begin{equation}
  %   p(\ubm{s}|\ubm{x})
  % \end{equation}
  The ``expected likelihood'' function:
  \begin{equation}\label{eq:em-q-funciton}
    \Qq(\bm{H}; \bm{H}^{\mathrm{old}}) = \EE_{p(\ubmr{s}{r}| \ubmr{x}{r}; \bm{H}^{\mathrm{old}})}\left[ \sum_{r=1}^{R}\log\,p(\ubmr{x}{r}, \ubmr{s}{r}; \bm{H}) \right]
  \end{equation}
\item M-step: the optimization step:
  \begin{equation}\label{eq:em-m-opt}
    \umax{\bm{H}} \Qq(\bm{H}; \bm{H}^{\mathrm{old}})
  \end{equation}
\end{itemize}

@@

\begin{algorithm}[H]
  \caption{Meta algorithm for HMM powered by Generative Models}
  \begin{algorithmic}[1]
    \STATE {\bfseries Input:}{
      Building $\bm{H}^{\mathrm{old}}, \bm{H} \in \Hh$ gives: \\
      $\bm{H}^{\mathrm{old}} = \{\Ss, \bm{q}^{\mathrm{old}}, A^{\mathrm{old}}, p(\bm{x}|s; \bm{\Phi}^{\mathrm{old}})\}$, $\bm{H} = \{\Ss, \bm{q}, A, p(\bm{x}|s; \bm{\Phi})\}$
    }, \\
    \STATE Initialize $\bm{H}$
    \STATE $\bm{H}^{\mathrm{old}} \gets \bm{H}$
    \FOR { $\bm{H}$ not converge}
    \STATE Sample a batch of data $\left\{ \ubmr{x}{r} \right\}_{r=1}^{R_b}$ from the dataset $\hat{p}(\ubm{x})$
    \STATE Compute $p(\smtr{s}{t}{r} | \ubmr{x}{r}; \bm{H}^{\mathrm{old}})$, $p(\smtr{s}{t}{r}, \smtr{s}{t+1}{r}| \ubmr{x}{r}; \bm{H}^{\mathrm{old}})$ by forward/backward algorithm;
    \STATE $\bm{q} \gets \uargmin{\bm{q}}\, \Qq(\bm{q}; \bm{H}^{\mathrm{old}})$ by \autoref{eq:update-initial-state-prob};
    \STATE $\bm{A} \gets \uargmin{\bm{A}}\Qq(\bm{A}; \bm{H}^{\mathrm{old}})$ by \autoref{eq:update-transition-prob};
    \STATE $\bm{\Phi} \gets \uargmin{\bm{\Phi}}\Qq(\bm{\Phi}; \bm{H}^{\mathrm{old}})$ by calling Algorithm \autoref{algo:opt-Phi-GenHMM} or \autoref{algo:opt-Phi-LatHMM};
    \STATE $\bm{H}^{\mathrm{old}} \gets \bm{H}$

@@

    % \node[enode] (z) at (0,0) {$\bm{z}\sim p(\bm{z})$};
    % \node[enode] (x) at (5.5,0){$\bm{x}\sim p(\bm{x}; \bm{\Phi})$};
    %               % \node at (5.2,-1) {$p(\bm{x};\bm{\Phi}) = \textstyle\sum_{k=1}^K \pi_k  p_k(\bm{x})$};
    % \node[nnode] (g1) at (2.6,1.8) {$\bm{g}_1$};
    % \node[nnode] (g2) at (2.6,0.5) {$\bm{g}_2$};
    % \node[nnode] (gk) at (2.6,-1.8) {$\bm{g}_K$};
    % \draw[dotted,line width=2pt] (2.6,-0.3) -- (2.6,-1.2);
    % \draw[->] (z) [in= 180, out =0] to (g1);
    % \draw[->] (z) [in= 180, out =0] to (g2);
    % \draw[->] (z) [in= 180, out =0] to (gk);
    % \filldraw[->] (3.7, 0.5)circle (2pt) -- node[above=0.2]{$\bm{s}\sim \bm{\pi}$} (x) ;
    %               % \draw[->] (3,-0.8) -- (3.5, -0.8);
    % \draw[->] (g1) -- (3.5,1.8);
    % \draw[->] (g2) -- (3.5, 0.5);
    % \draw[->] (gk) -- (3.5, -1.8);
  \end{tikzpicture}
  \caption{Source $s$ powered by generator mixed generative model (GenHMM).}
  \vspace{0.1cm}
\end{figure}

Base on the generative model assumption:

\begin{equation}
  p(\bm{x}| s; \bm{\Phi}_{s}) = \sum_{\kappa=1}^{K}\pi_{s, \kappa} p(\bm{x}| s, \kappa; \bm{\Phi}_{s, \kappa})
\end{equation}
where $\bm{\Phi}_s = \left\{ \bm{\Phi}_{s, \kappa}| \kappa = 1, 2, \cdots, K. \right\}$, and 
\begin{equation}
  \sum_{\kappa = 1}^{K} \pi_{s, \kappa}= 1
\end{equation}

@@

The help function should be revised to deal with the new latent variable $\kappa$ into:
\begin{align}\label{eq:obj-q-gen-mix}
  \Qq(\bm{\Phi}; \bm{H}^{\mathrm{old}}) & = \sum_{r=1}^{R} \sum_{t=1}^{{T}^{(r)}-1} \sum_{\smtr{s}{t}{r}=1}^{|\Ss|} \sum_{\smtr{\kappa}{t}{r}=1}^{K}p(\smtr{s}{t}{r}, \smtr{\kappa}{t}{r}| \ubmr{x}{r}; \bm{H}^{\mathrm{old}})  \log\,p(\smtr{\kappa}{t}{r}, \bmtr{x}{t}{r}| \smtr{s}{t}{r}; \bm{\Phi})\nonumber \\
  =& \sum_{r=1}^{R} \sum_{t=1}^{{T}^{(r)}-1} \sum_{\smtr{s}{t}{r}=1}^{|\Ss|}  \sum_{\smtr{\kappa}{t}{r}=1}^{K}p(\smtr{s}{t}{r}| \ubmr{x}{r}; \bm{H}^{\mathrm{old}})p(\smtr{\kappa}{t}{r}|\smtr{s}{t}{r}, \ubmr{x}{r}; \bm{H}^{\mathrm{old}}) \bigg[\log\,\pi_{\smtr{s}{t}{r},\smtr{\kappa}{t}{r}} \nonumber \\
                                        & + \log\, p(\bmtr{x}{t}{r} | \smtr{s}{t}{r}, \smtr{\kappa}{t}{r}; \bm{\Phi}) \bigg]
\end{align}
                                          %                                           \Nn\left( \bm{f}_{\smtr{s}{t}{r},\smtr{\kappa}{t}{r}}(\bm{x}); \bm{\mu}_{\smtr{s}{t}{r},\smtr{\kappa}{t}{r}}, \bm{C}_{\smtr{s}{t}{r},\smtr{\kappa}{t}{r}} \right) \log\,\bigg| \det\, \left( \pd{\bm{f}_{\smtr{s}{t}{r},\smtr{\kappa}{t}{r}}(\bmtr{x}{t}{r})}{\bmtr{x}{t}{r}} \right) \bigg| \bigg].
                                          %   \end{align}
In \autoref{eq:obj-q-gen-mix}, $p(s_t| \ubm{x}, \bm{H}^{\mathrm{old}})$ is computed by forward/backward algorithm. The posterior of $\kappa$ is:
\begin{align}\label{eq:kappa-posterior}
  p(\kappa| s, \ubm{x}; \bm{H}^{\mathrm{old}})
  &=  \frac{p(\kappa, \ubm{x}| s; \bm{H}^{\mathrm{old}})}{p(\ubm{x}| s,\bm{H}^{\mathrm{old}})} \nonumber \\
  & = \frac{\pi_{s, \kappa} p(\ubm{x}| s, \kappa, \bm{H}^{\mathrm{old}})}{\sum_{\kappa=1}^{K} \pi_{s, \kappa} p(\ubm{x}| s, \kappa,\bm{H}^{\mathrm{old}})} \nonumber \\
  & = \frac{\pi_{s, \kappa} p(\bm{x}| s, \kappa, \bm{H}^{\mathrm{old}})}{\sum_{\kappa=1}^{K}  \pi_{s, \kappa} p(\bm{x}| s, \kappa,\bm{H}^{\mathrm{old}})} 
  % & = \frac{\pi_{s_t, \kappa_t} \Nn\left( \bm{f}_{{s}_{t},{\kappa}_{t}}(\bm{x}); \bm{\mu}_{{s}_{t},{\kappa}_{t}}, \bm{C}_{{s}_{t},{\kappa}_{t}} \right) \bigg| \det\, \left( \pd{\bm{f}_{{s}_{t},{\kappa}_{t}}(\bmtr{x}{t}{r})}{\bmtr{x}{t}{r}} \right) \bigg| }{\sum_{\kappa_t=1}^{K}\pi_{s_t, \kappa_t} \Nn\left( \bm{f}_{{s}_{t},{\kappa}_{t}}(\bm{x}); \bm{\mu}_{{s}_{t},{\kappa}_{t}}, \bm{C}_{{s}_{t},{\kappa}_{t}} \right) \bigg| \det\, \left( \pd{\bm{f}_{{s}_{t},{\kappa}_{t}}(\bmtr{x}{t}{r})}{\bmtr{x}{t}{r}} \right) \bigg|} \nonumber \\
\end{align}
where the last equation is due to the fact that only $\bm{x}_t$ among sequence $\ubm{x}$ depends on $s_t, \kappa_t$.

@@

The latent prior for mixture of each source $s$ is obtained by solving the following problem:
\begin{align}
  \pi_{s, \kappa} & = \uargmax{pi_{s, \kappa}} \Qq(\bm{\Phi}; \bm{H}^{\mathrm{old}}) \\ \nonumber
                  & s.t. \, \sum_{\kappa=1}^{K} \pi_{s, \kappa}= 1
\end{align}
which gives the solution:
\begin{equation}\label{eq:mix-latent-parameter-solution}
  \pi_{s, \kappa} = \frac{\sum_{r=1}^{R} \sum_{t=1}^{{T}^{(r)}-1} p(\smtr{s}{t}{r} =s, \smtr{\kappa}{t}{r}=\kappa | \ubmr{x}{r}; \bm{H}^{\mathrm{old}}) }{\sum_{k =1}^{K}\sum_{r=1}^{R} \sum_{t=1}^{{T}^{(r)}-1} p(\smtr{s}{t}{r} =s, \smtr{\kappa}{t}{r}=k | \ubmr{x}{r}; \bm{H}^{\mathrm{old}}) }
\end{equation}
where $p(s, \kappa | \ubm{x}; \bm{H}^{\mathrm{old}}) = p(s| \ubm{x}; \bm{H}^{\mathrm{old}}) p(\kappa | s, \ubm{x}; \bm{H}^{\mathrm{old}})$ that can be computed by results of forward/backward and \autoref{eq:kappa-posterior}.

>>>>

Since model $\bm{H}$ contains hidden sequential variable $\ubm{s}$, we can not directly solve the maximum likelihood problem in \autoref{eq:ml-of-hmm}. We use expectation maximization (EM) to address the hidden variable problem by
\begin{itemize}
\item E-step: % the posterior probability of $\ubm{s}$:
  % \begin{equation}
  %   p(\ubm{s}|\ubm{x})
  % \end{equation}
  The ``expected likelihood'' function:
  \begin{equation}\label{eq:em-q-funciton}
    \Qq(\bm{H}; \bm{H}^{\mathrm{old}}) = \ \sum_{r=1}^{R}E_{p(\ubmr{s}{r}| \ubmr{x}{r}; \bm{H}^{\mathrm{old}})}\left[\log\,p(\ubmr{x}{r}, \ubmr{s}{r}; \bm{H}) \right]
  \end{equation}
\item M-step: the optimization step:
  \begin{equation}\label{eq:em-m-opt}
    \umax{\bm{H}} \Qq(\bm{H}; \bm{H}^{\mathrm{old}})
  \end{equation}
\end{itemize}

@@

\begin{algorithm}[H]
  \caption{Meta algorithm for HMM powered by Generative Models (GMHMM)}
  \begin{algorithmic}[1]
    \STATE {\bfseries Input:}{
      Building $\bm{H}^{\mathrm{old}}, \bm{H} \in \Hh$ gives: \\
      $\bm{H}^{\mathrm{old}} = \{\Ss, \bm{q}^{\mathrm{old}}, A^{\mathrm{old}}, p(\bm{x}|s; \bm{\Phi}^{\mathrm{old}})\}$, $\bm{H} = \{\Ss, \bm{q}, A, p(\bm{x}|s; \bm{\Phi})\}$
    }, \\
    \STATE Initialize $\bm{H}$
    \STATE $\bm{H}^{\mathrm{old}} \gets \bm{H}$
    \FOR { $\bm{H}$ not converge}
    \STATE Sample a batch of data $\left\{ \ubmr{x}{r} \right\}_{r=1}^{R_b}$ from the dataset $\hat{p}(\ubm{x})$
    \STATE Compute $p(\smtr{s}{t}{r} | \ubmr{x}{r}; \bm{H}^{\mathrm{old}})$, $p(\smtr{s}{t}{r}, \smtr{s}{t+1}{r}| \ubmr{x}{r}; \bm{H}^{\mathrm{old}})$ by forward/backward algorithm;
    \STATE $\bm{q} \gets \uargmin{\bm{q}}\, \Qq(\bm{q}; \bm{H}^{\mathrm{old}})$ by \autoref{eq:update-initial-state-prob};
    \STATE $\bm{A} \gets \uargmin{\bm{A}}\Qq(\bm{A}; \bm{H}^{\mathrm{old}})$ by \autoref{eq:update-transition-prob};
    \STATE $\bm{\Phi} \gets \uargmin{\bm{\Phi}}\Qq(\bm{\Phi}; \bm{H}^{\mathrm{old}})$ by calling Algorithm \autoref{algo:opt-Phi-GenHMM} or \autoref{algo:opt-Phi-LatHMM};
    \STATE $\bm{H}^{\mathrm{old}} \gets \bm{H}$

@@

    % \node[enode] (z) at (0,0) {$\bm{z}\sim p(\bm{z})$};
    % \node[enode] (x) at (5.5,0){$\bm{x}\sim p(\bm{x}; \bm{\Phi})$};
    %               % \node at (5.2,-1) {$p(\bm{x};\bm{\Phi}) = \textstyle\sum_{k=1}^K \pi_k  p_k(\bm{x})$};
    % \node[nnode] (g1) at (2.6,1.8) {$\bm{g}_1$};
    % \node[nnode] (g2) at (2.6,0.5) {$\bm{g}_2$};
    % \node[nnode] (gk) at (2.6,-1.8) {$\bm{g}_K$};
    % \draw[dotted,line width=2pt] (2.6,-0.3) -- (2.6,-1.2);
    % \draw[->] (z) [in= 180, out =0] to (g1);
    % \draw[->] (z) [in= 180, out =0] to (g2);
    % \draw[->] (z) [in= 180, out =0] to (gk);
    % \filldraw[->] (3.7, 0.5)circle (2pt) -- node[above=0.2]{$\bm{s}\sim \bm{\pi}$} (x) ;
    %               % \draw[->] (3,-0.8) -- (3.5, -0.8);
    % \draw[->] (g1) -- (3.5,1.8);
    % \draw[->] (g2) -- (3.5, 0.5);
    % \draw[->] (gk) -- (3.5, -1.8);
  \end{tikzpicture}
  \caption{Source $s$ powered by generator mixed generative model (GenHMM).}
  \vspace{0.1cm}
\end{figure}
In this subsection, we specify the GenHMM model and give the learning algorithm of this model.
Base on the generative model assumption, for each state $s$, the probabilistic generative model is
\begin{equation}
  p(\bm{x}| s; \bm{\Phi}_{s}) = \sum_{\kappa=1}^{K}\pi_{s, \kappa} p(\bm{x}| s, \kappa; \bm{\Phi}_{s, \kappa})
\end{equation}
where $\bm{\Phi}_s = \left\{ \pi_{s, \kappa}, \bm{\Phi}_{s, \kappa}| \kappa = 1, 2, \cdots, K. \right\}$, $\pi_{s, \kappa} = p(\kappa | \bm{x}, s; \bm{\Phi}_{s, \kappa})$ and 
\begin{equation}
  \sum_{\kappa = 1}^{K} \pi_{s, \kappa}= 1
\end{equation}

@@

The help function should be revised to deal with the new latent variable $\kappa$ into:
\begin{align}\label{eq:obj-q-gen-mix}
  \Qq(\bm{\Phi}; \bm{H}^{\mathrm{old}}) & = \sum_{r=1}^{R} \sum_{t=1}^{{T}^{(r)}-1} \sum_{\smtr{s}{t}{r}=1}^{|\Ss|} \sum_{\smtr{\kappa}{t}{r}=1}^{K}p(\smtr{s}{t}{r}, \smtr{\kappa}{t}{r}| \ubmr{x}{r}; \bm{H}^{\mathrm{old}})  \log\,p(\smtr{\kappa}{t}{r}, \bmtr{x}{t}{r}| \smtr{s}{t}{r}; \bm{\Phi}_{s})\nonumber \\
  =& \sum_{r=1}^{R} \sum_{t=1}^{{T}^{(r)}-1} \sum_{\smtr{s}{t}{r}=1}^{|\Ss|}  \sum_{\smtr{\kappa}{t}{r}=1}^{K}p(\smtr{s}{t}{r}| \ubmr{x}{r}; \bm{H}^{\mathrm{old}})p(\smtr{\kappa}{t}{r}|\smtr{s}{t}{r}, \ubmr{x}{r}; \bm{H}^{\mathrm{old}}) \bigg[\log\,\pi_{\smtr{s}{t}{r},\smtr{\kappa}{t}{r}} \nonumber \\
                                        & + \log\, p(\bmtr{x}{t}{r} | \smtr{s}{t}{r}, \smtr{\kappa}{t}{r}; \bm{\Phi}_s) \bigg]
\end{align}
                                          %                                           \Nn\left( \bm{f}_{\smtr{s}{t}{r},\smtr{\kappa}{t}{r}}(\bm{x}); \bm{\mu}_{\smtr{s}{t}{r},\smtr{\kappa}{t}{r}}, \bm{C}_{\smtr{s}{t}{r},\smtr{\kappa}{t}{r}} \right) \log\,\bigg| \det\, \left( \pd{\bm{f}_{\smtr{s}{t}{r},\smtr{\kappa}{t}{r}}(\bmtr{x}{t}{r})}{\bmtr{x}{t}{r}} \right) \bigg| \bigg].
                                          %   \end{align}
In \autoref{eq:obj-q-gen-mix}, $p(s_t| \ubm{x}, \bm{H}^{\mathrm{old}})$ is computed by forward/backward algorithm. The posterior of $\kappa$ is:
\begin{align}\label{eq:kappa-posterior}
  p(\kappa| s, \ubm{x}; \bm{H}^{\mathrm{old}})
  &=  \frac{p(\kappa, \ubm{x}| s; \bm{H}^{\mathrm{old}})}{p(\ubm{x}| s,\bm{H}^{\mathrm{old}})} \nonumber \\
  & = \frac{\pi_{s, \kappa} p(\ubm{x}| s, \kappa, \bm{H}^{\mathrm{old}})}{\sum_{\kappa=1}^{K} \pi_{s, \kappa} p(\ubm{x}| s, \kappa,\bm{H}^{\mathrm{old}})} \nonumber \\
  & = \frac{\pi_{s, \kappa} p(\bm{x}| s, \kappa, \bm{H}^{\mathrm{old}})}{\sum_{\kappa=1}^{K}  \pi_{s, \kappa} p(\bm{x}| s, \kappa,\bm{H}^{\mathrm{old}})} 
  % & = \frac{\pi_{s_t, \kappa_t} \Nn\left( \bm{f}_{{s}_{t},{\kappa}_{t}}(\bm{x}); \bm{\mu}_{{s}_{t},{\kappa}_{t}}, \bm{C}_{{s}_{t},{\kappa}_{t}} \right) \bigg| \det\, \left( \pd{\bm{f}_{{s}_{t},{\kappa}_{t}}(\bmtr{x}{t}{r})}{\bmtr{x}{t}{r}} \right) \bigg| }{\sum_{\kappa_t=1}^{K}\pi_{s_t, \kappa_t} \Nn\left( \bm{f}_{{s}_{t},{\kappa}_{t}}(\bm{x}); \bm{\mu}_{{s}_{t},{\kappa}_{t}}, \bm{C}_{{s}_{t},{\kappa}_{t}} \right) \bigg| \det\, \left( \pd{\bm{f}_{{s}_{t},{\kappa}_{t}}(\bmtr{x}{t}{r})}{\bmtr{x}{t}{r}} \right) \bigg|} \nonumber \\
\end{align}
where the last equation is due to the fact that only $\bm{x}_t$ among sequence $\ubm{x}$ depends on $s_t, \kappa_t$.

@@

The latent prior for mixture of each source $s$ is obtained by solving the following problem:
\begin{align}\label{opm:pi}
  \pi_{s, \kappa} & = \uargmax{\pi_{s, \kappa}} \Qq(\bm{\Phi}; \bm{H}^{\mathrm{old}}) \\ \nonumber
                  & s.t. \, \sum_{\kappa=1}^{K} \pi_{s, \kappa}= 1, \forall s = 1, 2, \cdots, |\Ss|. 
\end{align}

To solve problem \autoref{opm:pi}, we formulate its Lagrange function as
\begin{equation}
  \Ll = \Qq(\bm{\Phi}; \bm{H}^{\mathrm{old}}) + \sum_{\lambda_s} \lambda_s\left( 1-  \sum_{\kappa=1}^{K}\pi_{s, \kappa}  \right).
\end{equation}
Solving $\pd{\Ll}{\pi_{s, \kappa}} = 0$ gives
\begin{equation}
  \pi_{s,\kappa} = \frac{1}{\lambda_s}\sum_{r=1}^{R} \sum_{t=1}^{{T}^{(r)}} p(\smtr{s}{t}{r}=s, \smtr{\kappa}{t}{r} =\kappa| \ubmr{x}{r}; \bm{H}^{\mathrm{old}})
\end{equation}
With condition $\sum_{\kappa=1}^{K} \pi_{s, \kappa}= 1, \forall s = 1, 2, \cdots, |\Ss|$, we have
\begin{equation}
  \lambda_s = \sum_{\kappa=1}^{K}\sum_{r=1}^{R} \sum_{t=1}^{{T}^{(r)}} p(\smtr{s}{t}{r}=s, \smtr{\kappa}{t}{r} =\kappa | \ubmr{x}{r}; \bm{H}^{\mathrm{old}})
\end{equation}
Then the solution to \autoref{opm:pi} is
\begin{equation}\label{eq:mix-latent-parameter-solution}
  \pi_{s, \kappa} = \frac{\sum_{r=1}^{R} \sum_{t=1}^{{T}^{(r)}} p(\smtr{s}{t}{r} =s, \smtr{\kappa}{t}{r}=\kappa | \ubmr{x}{r}; \bm{H}^{\mathrm{old}}) }{\sum_{k =1}^{K}\sum_{r=1}^{R} \sum_{t=1}^{{T}^{(r)}-1} p(\smtr{s}{t}{r} =s, \smtr{\kappa}{t}{r}=k | \ubmr{x}{r}; \bm{H}^{\mathrm{old}}) }
\end{equation}
where $p(s, \kappa | \ubm{x}; \bm{H}^{\mathrm{old}}) = p(s| \ubm{x}; \bm{H}^{\mathrm{old}}) p(\kappa | s, \ubm{x}; \bm{H}^{\mathrm{old}})$ that can be computed by results of forward/backward and \autoref{eq:kappa-posterior}.

\textcolor{red}{I am going to put specific update formula for $\pi$ here....}
